State-space PTA

### 2023-06-16 docs/manuscript/paper.tex
response to Julian comments 1

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}
The detection of high frequency ($\sim 100$ Hz) gravitational waves (GWs) from coalescing black hole (BH) binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}. GWs from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2037$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper}, especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}. Detecting GWs from systems which evolve over even longer timescales, $\mathcal{O}$(years), has necessitated the development of novel astrophysical methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines. The foremost technique for the detection of GWs in this nano-Hz regime is a via timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{2021hgwa.bookE...4V}. The presence  of a nHz GW will influence the propagation of the pulsar radio beacon, leaving a characteristic impression on the received pulsar timing signal. By measuring the modulation of the pulsar signal in this way, one can effectively construct a detector with a baseline on the scale of parsecs. Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}. These previously disparate efforts have now been joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}. \newline

The primary target of PTA observations is the gravitational radiation emitted from the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R}, in addition to other potential GW sources such as cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}. The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. Previous efforts have mainly focused on the detection of this stochastic background via measuring the correlation in the pulsar timing residuals between any two pair of pulsars; the presence of a GW induces a characteristic correlation as a function of the angular separation between the pulsars - the Hellings-Downs curve \citep{Hellings}. However, no GW background has yet been detected \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A}. \newline

SMBHBs which are sufficiently massive and nearby may be individually resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. 
The stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}. However, as with the stochastic background, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Zhu2014PPTA,Babak2016,Arzoumanian2023}. The intrinsic pulsar timing noise - i.e.  random variations in the pulse arrival time - has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}. This pulsar timing noise has multiple theorized causes including microglitches \citep{Melatos2008}, glitch recovery \citep{Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Antonelli2023} and superfluid turbulence \citep{Melatos2014}. In order to mitigate the influence of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are known to be much more stable rotators with minimal timing noise. However, timing noise in MSPs could be a `latent' phenomenon that will emerge with the increased detector sensitivity and longer timing spans necessary to detect GWs \citep{Shannon2010}. In addition to achromatic timing noise there are also secondary, chromatic, noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

In this work we present a novel approach to PTA data analysis for individual, monochromatic, SMBHB GW sources. This method can self-consistently account for the intrinsic stochastic processes inherent to the pulsar. We formulate PTA analysis as a state-space problem and demonstrate how to track the state-space evolution optimally by using a Kalman filter. With optimal tracking of the intrinsic pulsar state, it is then possible to determine the presence of a GW signal in the pulsar data and infer the underlying source parameters. \newline

@@

>>>>

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}
The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R},  would result in the emission of nHz gravitational waves (GWs). Other potential GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}. The detection of these long-wavelength GWs has necessitated the development of novel astrophysical methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines. The foremost technique for the detection of GWs in this  low-frequency regime is a via timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{2021hgwa.bookE...4V}. The presence  of a nHz GW will influence the propagation of the pulsar radio beacon, leaving a characteristic impression on the received pulsar timing signal. By measuring the modulation of the pulsar signal in this way, one can effectively construct a detector with a baseline on the scale of parsecs. Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}. These previously disparate efforts have now been joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, as MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep{Hobbs_2019}. \newline

The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. Previous efforts have mainly focused on the detection of this stochastic background via measuring the correlation in the pulsar timing residuals between any two pair of pulsars; the presence of a GW induces a characteristic correlation as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}. However, no GW background has yet been detected \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A}. SMBHBs which are sufficiently massive and nearby may be individually resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. 
The stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}. However, as with the stochastic background, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Zhu2014PPTA,Babak2016,Arzoumanian2023}. \newline

The intrinsic pulsar timing noise - i.e.  random, unmodelled variations in the pulse arrival time due to irregularities in the rotation of the star - has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}. This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J}, magnetospheric state switching \citep{magneto1,Lyne2010L} and superfluid turbulence \citep{Link10.111,Melatos2014}. In order to mitigate the influence of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are known to be much more stable rotators with minimal timing noise. However, timing noise in MSPs could be a ``latent" phenomenon that will emerge with the increased detector sensitivity and longer timing spans necessary to detect GWs \citep{Shannon2010}. In addition to the temporally uncorrelated ``red" timing noise there are also secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

In this work we present a novel approach to PTA data analysis for individual, monochromatic, SMBHB GW sources that can self-consistently account for the intrinsic timing noise inherent to pulsar observations. We formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution by using a Kalman filter 
\citep{Kalman1,Meyers2021,Melatos2023}.  With the tracking of the intrinsic pulsar state established, it is then possible to determine the presence of a GW signal in the pulsar data and infer the underlying source parameters. \newline

@@

%
%The detection of high frequency ($\sim 100$ Hz) gravitational waves (GWs) from coalescing black hole (BH) binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}. GWs from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2037$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper}, especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}.

%
% Detecting GWs from systems which evolve over even longer timescales, $\mathcal{O}$(years), 
%
